---
title: "One-line Monte Carlo simulation"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
    mathjax: "default"
---


## 1. Sample Means

For each group, the sample mean is calculated as:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

## 2. Sample Variance and Standard Deviation

Sample variance:

$$
s^2 = \frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2
$$

Sample standard deviation:

$$
s = \sqrt{s^2}
$$

Let \( s_1, s_2 \) and \( n_1, n_2 \) be the standard deviations and sample sizes of group 1 and group 2.

## 3. Standard Error of the Difference Between Means

Assuming independent samples:

$$
SE = \sqrt{ \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} }
$$

## 4. Test Statistic (t)

The t-statistic is calculated as:

$$
t = \frac{ \bar{x}_1 - \bar{x}_2 }{SE}
$$

This t-statistic represents the standardized difference between the sample means under the assumption that the null hypothesis \( H_0: \mu_1 = \mu_2 \) is true.

## 5. Degrees of Freedom (Welch’s Approximation)

Because the two samples may have different variances, we use the Welch-Satterthwaite approximation for degrees of freedom:

$$
df = \frac{ \left( \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} \right)^2 }{
\frac{ \left( \frac{s_1^2}{n_1} \right)^2 }{n_1 - 1} +
\frac{ \left( \frac{s_2^2}{n_2} \right)^2 }{n_2 - 1}
}
$$

This value of \( df \) determines which **t-distribution** we reference when calculating the p-value.

## 6. p-Value

Given the calculated t-statistic and degrees of freedom, the p-value is computed as the probability of observing a t-value at least as extreme as the one obtained:

- For a two-sided test:

$$
p = 2 \cdot P(T_{df} > |t|)
$$

Where \( T_{df} \) denotes a t-distribution with the calculated degrees of freedom.




## The t-Distribution

The **Student's t-distribution** arises when standardizing the difference between a sample mean and a population mean, using the **sample standard deviation** instead of the population standard deviation. If \( X_1, \dots, X_n \sim \mathcal{N}(\mu, \sigma^2) \), then:

$$
t = \frac{\bar{X} - \mu}{S / \sqrt{n}} \sim t_{n - 1}
$$

where:

- \( \bar{X} \) is the sample mean,
- \( S \) is the sample standard deviation,
- \( n \) is the sample size,
- and the degrees of freedom are \( df = n - 1 \).

The t-distribution is derived from the ratio of a standard normal variable and the square root of a scaled chi-squared variable:

$$
t = \frac{Z}{\sqrt{V / \nu}} \quad \text{where } Z \sim \mathcal{N}(0,1),\; V \sim \chi^2_{\nu},\; Z \perp V
$$

Thus, the t-distribution with \( \nu \) degrees of freedom is defined by:

$$
t \sim \frac{Z}{\sqrt{V/\nu}} \sim t_{\nu}
$$

## Probability Density Function (PDF)

The probability density function of the t-distribution with \( \nu \) degrees of freedom is:

$$
f(t) = \frac{ \Gamma\left(\frac{\nu + 1}{2}\right) }
{ \sqrt{\nu \pi} \, \Gamma\left(\frac{\nu}{2}\right) }
\left(1 + \frac{t^2}{\nu} \right)^{ -\frac{\nu + 1}{2} }
$$

Where \( \Gamma(\cdot) \) is the gamma function, a generalization of the factorial:

$$
\Gamma(n) = (n - 1)!
$$

This function defines the shape of the t-distribution used to determine p-values for a given t-statistic.


# combined

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

$$
s^2 = \frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2
$$

$$
SE = \sqrt{ \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} }
$$

$$
t = \frac{ \bar{x}_1 - \bar{x}_2 }{SE}
$$

$$
df = \frac{ \left( \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} \right)^2 }{
\frac{ \left( \frac{s_1^2}{n_1} \right)^2 }{n_1 - 1} +
\frac{ \left( \frac{s_2^2}{n_2} \right)^2 }{n_2 - 1}
}
$$

$$
t = \frac{\bar{X} - \mu}{S / \sqrt{n}} \sim t_{n - 1}
$$
$$
f(t) = \frac{ \Gamma\left(\frac{\nu + 1}{2}\right) }
{ \sqrt{\nu \pi} \, \Gamma\left(\frac{\nu}{2}\right) }
\left(1 + \frac{t^2}{\nu} \right)^{ -\frac{\nu + 1}{2} }
$$

$$
\Gamma(n) = (n - 1)!
$$

$$
p = 2 \cdot P(T_{df} > |t|)
$$

# plots

```{r fig.height=3, fig.width=5}

library(ggplot2)
library(ggdist)  # for geom_dots
library(janitor)

# Example data
#set.seed(123)

df <- data.frame(
  Group = rep(c("Placebo", "Intervention"), each = 50),
  Score = round_half_up(c(rnorm(50, mean = 17, sd = 4), 
                          rnorm(50, mean = 15, sd = 4)), 0)
)

# Dot-only raincloud-style plot
ggplot(df, aes(x = Group, y = Score)) +
  ggdist::geom_dots(
    side = "right",          # only show dots on the left side
    scale = 0.8,            # controls spacing
    dotsize = 0.75,          # size of dots
    binwidth = 1          # controls vertical binning
  ) +
  scale_y_continuous(name = "Depression score", breaks = scales::breaks_pretty(n = 10),
                     limits = c(0, 35)) +
  coord_flip() +            # horizontal layout, like a raincloud
  theme_linedraw() +
  labs(title = "Sample 5 (N = 50 per group)")

ggsave("rct5.png",
       width = 5, 
       height = 3)

```



```{r}

mu1 <- 17
sigma1 <- 1

mu2 <- 15
sigma2 <- 1

ggplot(data.frame(x = c(12, 20)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = mu1, sd = sigma1), 
                aes(color = "Placebo"), size = 1.2) +
  stat_function(fun = dnorm, args = list(mean = mu2, sd = sigma2), 
                aes(color = "Intervention"), size = 1.2) +
  scale_color_manual(values = c("Placebo" = "steelblue", "Intervention" = "tomato")) +
  labs(title = "Population Distribution",
       x = "x", y = "Density", color = "Group") +
  theme_linedraw() +
  scale_x_continuous(name = "Depression score", 
                     breaks = scales::breaks_pretty(n = 10),
                     limits = c(12, 20))

```



```{r}

x_vals <- seq(0, 63, length.out = 500)

df_plot <- bind_rows(
  data.frame(
    x = x_vals,
    density = dnorm(x_vals, mean = 17, sd = 4),
    Group = "Placebo"
  ),
  data.frame(
    x = x_vals,
    density = dnorm(x_vals, mean = 15, sd = 4),
    Group = "Intervention"
  )
)

# Plot with shaded area under the curve
ggplot(df_plot, aes(x = x, y = density, fill = Group)) +
  geom_area(alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("Placebo" = "steelblue", "Intervention" = "tomato")) +
  labs(title = "Population Distribution", x = "x", y = "Density") +
  theme_linedraw() +
  scale_x_continuous(name = "Depression score", 
                     breaks = scales::breaks_pretty(n = 8),
                     limits = c(0, 35))

ggsave("pop.png",
       width = 5, 
       height = 3)

```

# PRNG

```{r}

library(janitor)


runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)


runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)


runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)

```


```{r}

library(janitor)

set.seed(42)
runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)

set.seed(42)
runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)

set.seed(42)
runif(n = 3, min = 1, max = 10) |> round_half_up(digits = 0)

```

# multivariate data

```{r fig.height=5, fig.width=5}

# Load required packages
library(ggplot2)

library(ggExtra)

# Generate multivariate normal data
# For example, 500 observations of two variables with correlation = 0.6
set.seed(123)  # for reproducibility

library(faux)

dat <- rnorm_multi(n = 500, 
                   vars = 2, 
                   varnames = c("X", "Y"),
                   mu = 0, 
                   sd = 1, 
                   r = 0.6)

p <- ggplot(dat, aes(x = X, y = Y)) +
  geom_point(alpha = 0.6) +
  theme_linedraw()

ggMarginal(p, type = "density", fill = "gray", alpha = 0.5)

ggsave("scatter.png",
       width = 5, 
       height = 5)

```





